---
title: "An Analysis of MPG by Transmission Type"
author: "Matt Moramarco"
date: "June 22, 2014"
output: pdf_document
---

## Executive Summary

This paper seeks to address how a vehicle's transmission (automatic or manual) impacts that vehicles MPG.

```{r echo = FALSE}
data(mtcars)
```
```{r}
boxplot(mpg ~ am, data = mtcars, xlab = "Transmission", ylab = "MPG", main = "MPG by Transmission")
```

Given the above plot, it would seem that manual transmissions (labeled '1' above) have a higher MPG than automatic transmissions (labeled '0' above). 

The selected model for this analysis incorporates Transmission, Weight, and the interaction between the two as input variables upon on the predicted variable of MPG.

```{r}
fit3 <- lm(mpg ~ factor(am) * wt, data = mtcars)
fit3$coef
```

The positive coeffecient of the manual transmission (14.878) indicates the manual transmissions have a positive effect on MPG (all else being equal). However, this is complicated by the interaction of Weight and Transmission. For each thousand pounds of weight, a manual transmission can expect an additional reduction in MPG in the amount of -5.298 (note that this is in addition to the -3.786 decrease in MPG per thousand pounds that affects both manual and automatic transmissions.

Thus, this model indicates that a manual transmission is more efficient up to a specific weight: 2,808 pounds. At this point, an automatic transmission is more efficient in terms of MPG

```{r}
plot(predict(fit3,mtcars) , mtcars$mpg, xlab = "Predicted MPG", ylab = "Actual MPG")
abline(a=0, b = 1, col = "red")
```

## Appendix

We start by first looking at the data:

```{r results = 'hide'}
head(mtcars)
str(mtcars)
summary(mtcars)
```

The variable 'am' indicates the cars transmission type where 0 indicates automatic and 1 indicates manual.

### Model 1:

The first model, below, simply conducts a linear regression using mpg as the outcome variable and a factorized tranmission type as the input variable. This model produces a low R-squared indicating that it does not adequately explain the variance.

```{r}
fit <- lm(mpg ~ factor(am), data = mtcars)
fit$coef
```

### Model 2:

The second model builds upon the first but includes all available variables as input variables. While this model's R-squared is much higher than Model 1, explaining a larger portion of the variance, the complexity of the model (the inclusion of all the variables) is inefficient.

```{r}
fit2 <- lm(mpg ~ factor(am) + ., data = mtcars)
fit2$coef
```

We then look at pairwise plots for all variables against all other variables to gain a better understanding of how variables may interact with mpg.

``` {r}
pairs(mtcars, panel = panel.smooth)
```

Transmission, Weight and MPG are isolated in the below pairwise plot as these appear to have the strongest interactions.

```{r}
pairs(mtcars[c("am","wt","mpg")], panel = panel.smooth)
```

### Model 3:

Using the information from the above, we now isolate and use Transmission, Weight and the interaction between the two as inputs for our third and final model. This model produces a higher R-squared than Model 2 and is simpler to both explain and understand.

```{r}
summary(fit3)
par(mfrow = c(2, 2))
plot(fit3)
```